---
title: "01_DADA2"
author: "Maya Craig"
date: "`r Sys.Date()`"
output: html_document:
  code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/01_DADA2/") # send any figure output to this folder
```


## Set seed 
```{r}
set.seed(4750)
```


```{r}
# What time did we start running this script? 
start_time <- Sys.time()
start_time
```

# Goals of this file 

1. Use raw fastq files and generate quality plots to assess quality of reads.
2. Filter and trim out bad sequences and bases from our sequencing files. 
3. Write out fastq files with high quality sequences. 
4. Evaluate the quality from our filter and trim. 
5. Infer Errors on forward and reverse reads individually.
6. Identified ASVs on forward and reverse reads separately, using the error model.  
7. Merge forward and reverse ASVs into "contiguous ASVs".  
8. Generate the ASV count table. (`otu_table` input for phyloseq.). 


# Load packages 
```{r}
# Efficient package loading with pacman 
pacman::p_load(tidyverse, BiocManager, devtools, dada2, 
               phyloseq, patchwork, DT, iNEXT, vegan,
               install = FALSE)
```


# Load data 
```{r}
# Set the raw fastq path to the raw sequencing files 
# Path to the fastq files 
raw_fastqs_path <- "data/01_raw_gzipped_fastqs"
raw_fastqs_path
head(list.files(raw_fastqs_path))
str(list.files(raw_fastqs_path)) #how many files 
```

# Create vector of forward reads 
```{r}
forward_reads <- list.files(raw_fastqs_path, pattern = "_1.fastq.gz", full.names = TRUE)  
head(forward_reads)  
```

# Create vector of reverse reads
```{r}
reverse_reads <- list.files(raw_fastqs_path, pattern = "_2.fastq.gz", full.names = TRUE)  
head(reverse_reads)  
```

# Asses Read Quality 

# Evaluate raw read sequence quality 

- Examining the quality of reads before trimming 

```{r}
# Randomly select 12 samples from dataset to evaluate 
random_samples <- sample(1:length(reverse_reads), size = 12)
random_samples

# Calculate and plot quality of these two samples
forward_filteredQual_plot_12 <- plotQualityProfile(forward_reads[random_samples]) + 
  labs(title = "Forward Read: Raw Quality")
```

