---
title: "01_DADA2"
author: "Maya Craig"
date: "`r Sys.Date()`"
output: html_document:
  code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.path = "../figures/01_DADA2/") # send any figure output to this folder
```


## Set seed 
```{r}
set.seed(4750)
```


```{r}
# What time did we start running this script? 
start_time <- Sys.time()
start_time
```

# Goals of this file 

1. Use raw fastq files and generate quality plots to assess quality of reads.
2. Filter and trim out bad sequences and bases from our sequencing files. 
3. Write out fastq files with high quality sequences. 
4. Evaluate the quality from our filter and trim. 
5. Infer Errors on forward and reverse reads individually.
6. Identified ASVs on forward and reverse reads separately, using the error model.  
7. Merge forward and reverse ASVs into "contiguous ASVs".  
8. Generate the ASV count table. (`otu_table` input for phyloseq.). 


# Load packages 
```{r load-packages}
# Efficient package loading with pacman 
pacman::p_load(tidyverse, BiocManager, devtools, dada2, 
               phyloseq, patchwork, DT, iNEXT, vegan,
               install = FALSE)
library(ggplot2)
```


# Load data 
```{r load-data}
# Set the raw fastq path to the raw sequencing files 
# Path to the fastq files 
raw_fastqs_path <- "data/01_raw_gzipped_fastqs"
raw_fastqs_path
head(list.files(raw_fastqs_path))
str(list.files(raw_fastqs_path)) #how many files 
```

# Create vector of forward reads 
```{r forward-reads-vector}
forward_reads <- list.files(raw_fastqs_path, pattern = "_1.fastq.gz", full.names = TRUE)  
head(forward_reads)  
```

# Create vector of reverse reads
```{r reverse-reads-vector}
reverse_reads <- list.files(raw_fastqs_path, pattern = "_2.fastq.gz", full.names = TRUE)  
head(reverse_reads)  
```

# Asses Read Quality 

# Evaluate raw read sequence quality 

- Examining the quality of reads before trimming 

```{r raw-aggregate-plot}
# Randomly select 12 samples from dataset to evaluate 
random_samples <- sample(1:length(reverse_reads), size = 12)
random_samples

# Calculate and plot quality of these two samples
forward_filteredQual_plot_12 <- plotQualityProfile(forward_reads[random_samples]) + 
  labs(title = "Forward Read: Raw Quality")

reverse_filteredQual_plot_12 <- plotQualityProfile(reverse_reads[random_samples]) + 
  labs(title = "Reverse Read: Raw Quality")

preQC_aggregate_plot <- forward_filteredQual_plot_12 + reverse_filteredQual_plot_12

preQC_aggregate_plot
```

## Prepare a placeholder for filtered reads
```{r prep-filtered-sequences}
# vector of our samples, extract sample name from files 
samples <- sapply(strsplit(basename(forward_reads), "_"), `[`,1) 
head(samples)

# Place filtered reads into filtered_fastqs_path
filtered_fastqs_path <- "data/01_DADA2/02_filtered_fastqs"
filtered_fastqs_path

# create 2 variables: filtered_F, filtered_R
filtered_forward_reads <- 
  file.path(filtered_fastqs_path, paste0(samples, "_1_filtered.fastq.gz"))
length(filtered_forward_reads)

# reverse reads
filtered_reverse_reads <- 
  file.path(filtered_fastqs_path, paste0(samples, "_2_filtered.fastq.gz"))
head(filtered_reverse_reads)
```

# Filter and Trim Reads

Parameters of filter and trim **DEPEND ON THE DATASET**. The things to keep in mind are:  
- The library preparation: *Are the primers included in the sequence? If so, they need to be trimmed out in this step*.  
- What do the above quality profiles of the reads look like? *If they are lower quality, it is highly recommended to use `maxEE = c(1,1)`.*  
- Do the reads dip suddenly in their quality? If so, explore `trimLeft` and `truncLen`

- `maxEE` is a quality filtering threshold applied to expected errors. Here, if there's 2 expected errors. It's ok. But more than 2. Throw away the sequence. Two values, first is for forward reads; second is for reverse reads.  
- `trimLeft` can be used to remove the beginning bases of a read (e.g. to trim out primers!) 
- `truncLen` can be used to trim your sequences after a specific base pair when the quality gets lower. Though, please note that this will shorten the ASVs! For example, this can be used when the quality of the sequence suddenly gets lower, or clearly is typically lower. So, if the quality of the read drops below a phred score of 25 (on the y-axis of the plotQualityProfile above, which indicates ~99.5% confidence per base).  
- `maxN` the number of N bases. Here, using ASVs, we should ALWAYS remove all Ns from the data.  

```{r filter-and-trim}
# Assign a vector to filtered reads 
# trim out poor bases, first 3 bps on F reads
# write out filtered fastq files 
# Therefore, we do not need to trim the primers, because they were not sequenced
filtered_reads <- 
  filterAndTrim(fwd = forward_reads, filt = filtered_forward_reads,
              rev = reverse_reads, filt.rev = filtered_reverse_reads,
              maxN = 0, maxEE = c(2,2), trimLeft = 0,
              truncQ = 2, rm.phix = TRUE, compress = TRUE, multithread = TRUE)
?filterAndTrim
```

# Plot the 12 random samples after QC
```{r}
forward_filteredQual_plot_12 <- 
  plotQualityProfile(filtered_forward_reads[random_samples]) + 
  labs(title = "Trimmed Forward Read Quality")

reverse_filteredQual_plot_12 <- 
  plotQualityProfile(filtered_reverse_reads[random_samples]) + 
  labs(title = "Trimmed Reverse Read Quality")

# Put the two plots together 
#forward_filteredQual_plot_12 + reverse_filteredQual_plot_12
```

## Aggregated Trimmed Plots 
```{r aggregated-plots}
# Forward reads
forward_postQC_plot <- 
  plotQualityProfile(filtered_forward_reads, aggregate = TRUE) + 
  labs(title = "Forward Post-QC")

# reverse reads
reverse_postQC_plot <- 
  plotQualityProfile(filtered_reverse_reads, aggregate = TRUE) + 
  labs(title = "Reverse Post-QC")

postQC_aggregate_plot <- 
  forward_postQC_plot + reverse_postQC_plot

# Show the plot
#postQC_aggregate_plot
```

```{r}
# Make output into dataframe 
filtered_df <- as.data.frame(filtered_reads)
head(filtered_df)


# calculate some stats 
filtered_df %>%
  reframe(median_reads_in = median(reads.in),
          median_reads_out = median(reads.out),
          median_percent_retained = (median(reads.out)/median(reads.in)))
```

